### 2025-08-10 factor-analysis.qmd
20250806 - duplicate identifiers
20250809 - label functions

#### Specify Variables {#sec-pcaPrepareDataSpecifyVars}

@@

#### Standardize Variables {#sec-pcaPrepareDataStandardize}

@@

We generated the scree plot in @fig-efaScreePlot1 using the `fa.parallel()` function of the `psych` package [@R-psych].
The optimal coordinates and the acceleration factor attempt to operationalize the Cattell scree test: i.e., the "elbow" of the scree plot [@Ruscio2012].
The optimal coordinators factor is quantified using a series of linear equations to determine whether observed eigenvalues exceed the predicted values.
The acceleration factor is quantified using the acceleration of the curve, that is, the second derivative.
The Kaiser-Guttman rule states to keep principal components whose eigenvalues are greater than 1.
However, for [exploratory factor analysis](#efa) (as opposed to [PCA](#pca)), the criterion is to keep the factors whose eigenvalues are greater than zero (i.e., not the factors whose eigenvalues are greater than 1) [@Dinno2014].

@@

We generated the scree plot in @fig-efaScreePlot2 using the `nScree()` and `plotnScree()` functions of the `nFactors` package [@R-nFactors].

@@

We generated the very simple structure (VSS) plots in Figures [-@fig-efaVSSPlot1] and [-@fig-efaVSSPlot2] using the `vss()` and `nfactors()` functions of the `psych` package [@R-psych].
In addition to VSS plots, the output also provides additional criteria by which to determine the optimal number of factors, each for which lower values are better, including the Velicer minimum average partial (MAP) test, the Bayesian information criterion (BIC), the sample size-adjusted BIC (SABIC), and the root mean square error of approximation (RMSEA).
Depending on the criterion, the optimal number of factors extracted varies between 2 and 8 factors.

@@

We fit EFA models using the `efa()` function of the `lavaan` package [@Rosseel2012_packages; @R-lavaan].

@@

A path diagram of the three-factor EFA model in @fig-efaPathDiagram-3factor was created using the `lavaanPlot()` function of the `lavaanPlot` package [@R-lavaanPlot].

@@

To fit the ESEM model, we use the `sem()` function of the `lavaan` package [@Rosseel2012_packages; @R-lavaan].

@@

To make the plot interactive for editing, you can use the `plot_lavaan()` function of the `lavaangui` package [@R-lavaangui; @Karch2025_packages]:

@@

We fit CFA models using the `cfa()` function of the `lavaan` package [@Rosseel2012_packages; @R-lavaan].
We compare one-, two-, and three-factor CFA models.

>>>>

#### Specify Variables {#sec-factorAnalysisPrepareDataSpecifyVars}

@@

#### Standardize Variables {#sec-factorAnalysisPrepareDataStandardize}

@@

We generated the scree plot in @fig-efaScreePlot1 using the `psych::fa.parallel()` function of the `psych` package [@R-psych].
The optimal coordinates and the acceleration factor attempt to operationalize the Cattell scree test: i.e., the "elbow" of the scree plot [@Ruscio2012].
The optimal coordinators factor is quantified using a series of linear equations to determine whether observed eigenvalues exceed the predicted values.
The acceleration factor is quantified using the acceleration of the curve, that is, the second derivative.
The Kaiser-Guttman rule states to keep principal components whose eigenvalues are greater than 1.
However, for [exploratory factor analysis](#efa) (as opposed to [PCA](#pca)), the criterion is to keep the factors whose eigenvalues are greater than zero (i.e., not the factors whose eigenvalues are greater than 1) [@Dinno2014].

@@

We generated the scree plot in @fig-efaScreePlot2 using the `nFactors::nScree()` and `nFactors::plotnScree()` functions of the `nFactors` package [@R-nFactors].

@@

We generated the very simple structure (VSS) plots in Figures [-@fig-efaVSSPlot1] and [-@fig-efaVSSPlot2] using the `psych::vss()` and `psych::nfactors()` functions of the `psych` package [@R-psych].
In addition to VSS plots, the output also provides additional criteria by which to determine the optimal number of factors, each for which lower values are better, including the Velicer minimum average partial (MAP) test, the Bayesian information criterion (BIC), the sample size-adjusted BIC (SABIC), and the root mean square error of approximation (RMSEA).
Depending on the criterion, the optimal number of factors extracted varies between 2 and 8 factors.

@@

We fit EFA models using the `lavaan::efa()` function of the `lavaan` package [@Rosseel2012_packages; @R-lavaan].

@@

A path diagram of the three-factor EFA model in @fig-efaPathDiagram-3factor was created using the `lavaanPlot::lavaanPlot()` function of the `lavaanPlot` package [@R-lavaanPlot].

@@

To fit the ESEM model, we use the `lavaan::sem()` function of the `lavaan` package [@Rosseel2012_packages; @R-lavaan].

@@

To make the plot interactive for editing, you can use the `lavaangui::plot_lavaan()` function of the `lavaangui` package [@R-lavaangui; @Karch2025_packages]:

@@

We fit CFA models using the `lavaan::cfa()` function of the `lavaan` package [@Rosseel2012_packages; @R-lavaan].
We compare one-, two-, and three-factor CFA models.

### 2025-08-10 getting-started.qmd
20250809 - label functions

   install.packages(c(
     "petersenlab","remotes","knitr","rmarkdown","tidyverse","nflverse",
     "tidymodels","easystats","broom","broom.mixed","psych","downlit","xml2",
     "gsisdecoder","progressr","DescTools","pwr","pwrss","WebPower","XICOR",
     "dagitty","ggdag","ggtext","gghighlight","ggExtra","grid","patchwork",
     "pROC", "lme4","lmerTest","MuMIn","emmeans","pbkrtest","sjstats",
     "AICcmodavg","rstan","brms","tidybayes","bbmle","fitdistrplus","sn",
     "mclust","magrittr","viridis","viridisLite","msir","plotly","webshot2",
     "quantmod","fPortfolio","NMOF","nFactors","xts","zoo","forecast","stringi",
     "parallelly","doParallel","missRanger","ggridges","powerjoin","caret",
     "LongituRF","gpboost","mgcv","rms","car","lavaan","lavaanPlot","lavaangui",
     "mice","miceadds","interactions","robustbase","ordinal","MASS",
     "data.table","future","future.apply","SimDesign","domir","GGally"))
   ```
1. Some necessary packages, including the [`ffanalytics` package](https://github.com/FantasyFootballAnalytics/ffanalytics) [@R-ffanalytics], are hosted in GitHub (and are not hosted on the Comprehensive R Archive Network [CRAN]) and thus need to be installed using the following code (after installing the `remotes` package [@R-remotes] above)[^petersenlabPackageGitHub]:
   ```{r}
   #| eval: false

@@

You can install `R` packages using the following syntax:

@@

You can load a given library using the following syntax:

@@

Below, we provide examples for how to learn about and use functions and arguments, by using the `seq()` function as an example.
The `seq()` function creates a sequence of numbers.
To learn about the `seq()` function, which creates a sequence of numbers, you can execute the following command:

@@

This is what the documentation shows for the `seq()` function in the `Usage` section:

@@

Based on this information, we know that the `seq()` function takes the following arguments:

@@

A vector is a series of elements that can be numeric or character.
Character elements should be specified in quotes.
A vector has one dimension (length).
To create a vector, use the `c()` function to combine elements into a vector ("c" stands for combine).
And, we use the assignment operator (`<-`) to assign the vector to an object named `exampleVector`, so we can access it later.
Anything on the right side of an assignment operator gets assigned to the object name on the left side of the assignment operator.

@@

A data frame has two dimensions: rows and columns.
Here is an example of creating a data frame, while using the assignment operator (`<-`) to assign the data frame to an object so we can access it later:

@@

fantasyPoints_weekly <- expand.grid(
  ID = 1:12,
  season = c(2022, 2023),
  week = 1:17
)

set.seed(52242)
fantasyPoints_weekly$fantasyPoints <- sample(
  0:35,
  size = nrow(fantasyPoints_weekly),
  replace = TRUE
)
```

@@

A list can store multiple data frames in one object:

@@

Here is how you load a `.RData` file using a *relative path* (i.e., a path relative to the working directory, where the working directory is represented by a period):

@@

The direction of the slashes matters—you should use forward slashes (not backslashes)!
To determine where you working directory is, you can type:

@@

To change your working directory, you can type:

@@

nfl_players <- read.csv("C:/Users/myusername/nfl_players.RData")
```

Here is how you load a `.csv` file:

@@

Here is how you save a `.RData` file using a *relative path* (i.e., relative to your working directory):

@@

Here is how you save a `.csv` file:

@@

To see the names of variables in a data frame, use the following syntax:

@@

The [`petersenlab`](https://cran.r-project.org/web/packages/petersenlab/index.html) package [@R-petersenlab] has a convenience function, `%ni%`, for the logical operator of determining whether an element is *not* in a value of another vector:

@@

@@

The operator, `which()`, can be used to obtain the TRUE indices of an object, which is useful for [subsetting](#sec-subset).

@@

To apply conditional operations to a vector, we can use the `ifelse()` function.

@@

In base `R`, you can perform piping with the `|>` expression.
In `tidyverse` you can perform piping with the `%>%` expression.

@@

@@

@@

@@

To view data, use the following syntax:

@@

To view only the first six rows (if a data frame) or elements (if a vector), use the following syntax:

@@

@@

Number of rows and columns:

@@

Number of rows:

@@

Number of columns:

@@

@@

You can determine how many rows in the data frame have no missing data using the following syntax:

@@

Recode multiple variables:

@@

@@

Using a vector of variable names:

@@

One variable:

@@

It is important not to have rows with duplicate values on the keys.
For instance, if there is more than one row with the same `player_id` in each object (or multiple rows in object 2 with the same combination of `player_id`, `season`, and `week`), then each row with that `player_id` in object 1 gets paired with each row with that `player_id` in object 2.
The many possible combinations can lead to the resulting object greatly expanding in terms of the number of rows.
Thus, you want the keys to uniquely identify each row.
In the example below, `player` is present in each object, so we can merge by `player`; however, each object has multiple rows with the same player.
For example, `mergeExample1A` has three rows for `player` A; `mergeExample1B` has two rows for `player` A.
Thus, when we merge them, the resulting object has many more rows than each respective object (even though neither object has players that the other object does not).

@@

Note: if the two objects include variables with the same name (apart from the keys), `R` will not know how you want each to appear in the merged object.
So, it will add a suffix (e.g., `.x`, `.y`) to each common variable to indicate which object (i.e., object `x` or object `y`) the variable came from, where object `x` is the first object—i.e., the object to which object `y` (the second object) is merged.
In general, apart from the keys, you should not include variables with the same name in two objects to be merged.
To prevent this, either remove or rename the shared variable in one of the objects, or include the shared variable as a key.
However, as described above, you should include it as a key ***only*** if you want to use its values to align the rows from each object.
Below is an example of merging two objects with the same variable name (i.e., points) that is not used as a key.

@@

When two objects are merged that have different formats, the resulting data object inherits the format of the data object that has more levels of nesting.
For instance, consider that you want to merge two objects, object A and object B.
Object A is in `player` form and object B is in `player`-`season`-`week` form.
When you merge them, the resulting data object will be in `player`-`season`-`week` form.

@@

A full outer join includes all rows in `x` **or** `y`.
It returns columns from `x` and `y`.
Here is how to merge two data frames using a full outer join (i.e., "full join"):

@@

A left outer join includes all rows in `x`.
It returns columns from `x` and `y`.
Here is how to merge two data frames using a left outer join ("left join"):

@@

A right outer join includes all rows in `y`.
It returns columns from `x` and `y`.
Here is how to merge two data frames using a right outer join ("right join"):

@@

An inner join includes all rows that are in **both** `x` **and** `y`.
An inner join will return one row of `x` for each matching row of `y`, and can duplicate values of records on either side (left or right) if `x` and `y` have more than one matching record.
It returns columns from `x` and `y`.
Here is how to merge two data frames using an inner join:

@@

A semi join is a filter.
A left semi join returns all rows from `x` **with** a match in `y`.
That is, it filters out records from `x` that are not in `y`.
Unlike an [inner join](#innerJoin), a left semi join will never duplicate rows of `x`, and it includes columns from only `x` (not from `y`).
Here is how to merge two data frames using a left semi join:

@@

An anti join is a filter.
A left anti join returns all rows from `x` **without** a match in `y`.
That is, it filters out records from `x` that are in `y`.
It returns columns from only `x` (not from `y`).
Here is how to merge two data frames using a left anti join:

@@

A cross join combines each row in `x` with each row in `y`.

@@

Below, we transform the data from long form to wide form.
The long data are in "`player`-`season`-`week` form".
We widen the data by two variables (`season` and `week`), using `tidyverse`, so that the data are now in "`player` form" (where each row is uniquely identified by the `ID` variable):

@@

Conversely, we can also restructure data from wide to long.
Here are the data in long form, after they have been transformed from wide form using `tidyverse`:

>>>>

   install.packages(c(
     "petersenlab","remotes","knitr","rmarkdown","tidyverse","nflverse",
     "tidymodels","easystats","broom","broom.mixed","psych","downlit","xml2",
     "gsisdecoder","progressr","DescTools","pwr","pwrss","WebPower","XICOR",
     "dagitty","ggdag","ggtext","gghighlight","ggExtra","grid","patchwork",
     "pROC", "lme4","lmerTest","MuMIn","emmeans","pbkrtest","sjstats",
     "AICcmodavg","rstan","brms","tidybayes","bbmle","fitdistrplus","sn",
     "mclust","magrittr","viridis","viridisLite","msir","plotly","webshot2",
     "quantmod","fPortfolio","NMOF","nFactors","xts","zoo","forecast","stringi",
     "parallelly","doParallel","missRanger","ggridges","powerjoin","caret",
     "LongituRF","gpboost","mgcv","rms","car","lavaan","lavaanPlot","lavaangui",
     "mice","miceadds","interactions","robustbase","ordinal","MASS",
     "data.table","future","future.apply","SimDesign","domir","GGally","Rglpk",
     "TTR"))
   ```
1. Some necessary packages, including the [`ffanalytics` package](https://github.com/FantasyFootballAnalytics/ffanalytics) [@R-ffanalytics], are hosted in GitHub (and are not hosted on the Comprehensive R Archive Network [CRAN]) and thus need to be installed using the following code (after installing the `remotes` package [@R-remotes] above)[^petersenlabPackageGitHub]:
   ```{r}
   #| eval: false

@@

You can install `R` packages using the `utils::install.packages()` function with the following syntax:

@@

You can load a given library using the `base::library()` function with the following syntax:

@@

Below, we provide examples for how to learn about and use functions and arguments, by using the `base::seq()` function as an example.
The `base::seq()` function creates a sequence of numbers.
To learn about the `base::seq()` function, which creates a sequence of numbers, you can execute the following command:

@@

This is what the documentation shows for the `base::seq()` function in the `Usage` section:

@@

Based on this information, we know that the `base::seq()` function takes the following arguments:

@@

A vector is a series of elements that can be numeric or character.
Character elements should be specified in quotes.
A vector has one dimension (length).
To create a vector, use the `base::c()` function to combine elements into a vector ("c" stands for combine).
And, we use the assignment operator (`<-`) to assign the vector to an object named `exampleVector`, so we can access it later.
Anything on the right side of an assignment operator gets assigned to the object name on the left side of the assignment operator.

@@

A data frame has two dimensions: rows and columns.
Here is an example of creating a data frame using the `base::data.frame()` function, while using the assignment operator (`<-`) to assign the data frame to an object so we can access it later:

@@

```

We can also create a data frame from all combinations of variables, using the `base::expand.grid()` function:

```{r}
fantasyPoints_weekly <- expand.grid(
  ID = 1:12,
  season = c(2022, 2023),
  week = 1:17
)
```

We can create a vector that is randomly sampled from values using the `base::sample()` function:

```{r}
set.seed(52242)
fantasyPoints_weekly$fantasyPoints <- sample(
  0:35,
  size = nrow(fantasyPoints_weekly),
  replace = TRUE
)
```

@@

A list can store multiple data frames in one object using the `base::list()` function:

@@

Here is how you load a `.RData` file using a *relative path* (i.e., a path relative to the working directory, where the working directory is represented by a period) using the `base::load()` function:

@@

The direction of the slashes matters—you should use forward slashes (not backslashes)!
To determine where you working directory is, you can use the `base::getwd()` function:

@@

To change your working directory, you can use the `base::setwd()` function:

@@

load("C:/Users/myusername/nfl_players.RData")
```

Here is how you load a `.csv` file using the `utils::read.csv()` function:

@@

Here is how you save a `.RData` file using a *relative path* (i.e., relative to your working directory) using the `base::save()` function:

@@

Here is how you save a `.csv` file using the `utils::write.csv()` function:

@@

To see the names of variables in a data frame, use the `base::names()` function:

@@

The [`petersenlab`](https://cran.r-project.org/web/packages/petersenlab/index.html) package [@R-petersenlab] has a convenience function, `petersenlab::\`%ni%\``, for the logical operator of determining whether an element is *not* in a value of another vector:

@@

`base::is.na()`

@@

The operator, `base::which()`, can be used to obtain the TRUE indices of an object, which is useful for [subsetting](#sec-subset).

@@

To apply conditional operations to a vector, we can use the `base::ifelse()` function.

@@

In base `R`, you can perform piping with the `base::\`|>\`` expression.
In `tidyverse`, you can perform piping with the `magrittr::\`%>%\`` expression.

@@

In `tidyverse`, you cans select particular columns using the `dplyr::select()` function.

@@

In `tidyverse`, you cans select particular columns using the `dplyr::filter()` function.

@@

In `tidyverse`, you cans select particular rows and columns using the `dplyr::filter()` and `dplyr::select()` functions, respectively.

@@

To view data, use the `utils::View()` function:

@@

To view only the first six rows (if a data frame) or elements (if a vector), use the `utils::head()` function:

@@

`utils::str()`

@@

Number of rows and columns (`base::dim()`):

@@

Number of rows (`base::nrow()`):

@@

Number of columns (`base::ncol()`):

@@

`base::length()`

@@

You can determine how many rows in the data frame have no missing data using the following syntax and the `stats::na.omit()` function:

@@

Recode multiple variables using the `dplyr::mutate()` and `dplyr::case_match()` functions:

@@

`dplyr::rename()`

@@

Using a vector of variable names with the `dplyr::rename_with()` function:

@@

One variable (`base::factor()`; `base::as.numeric()`; `base::as.integer()`; `base::as.character()`):

@@

It is important not to have rows with duplicate values on the keys.
For instance, if there is more than one row with the same `player_id` in each object (or multiple rows in object 2 with the same combination of `player_id`, `season`, and `week`), then each row with that `player_id` in object 1 gets paired with each row with that `player_id` in object 2.
The many possible combinations can lead to the resulting object greatly expanding in terms of the number of rows.
Thus, you want the keys to uniquely identify each row.
In the example below, `player` is present in each object, so we can merge by `player`; however, each object has multiple rows with the same player.
For example, `mergeExample1A` has three rows for `player` A; `mergeExample1B` has two rows for `player` A.
Thus, when we merge them, the resulting object has many more rows than each respective object (even though neither object has players that the other object does not).
Below, we merge data using the `dplyr::full_join()` function:

@@

Note: if the two objects include variables with the same name (apart from the keys), `R` will not know how you want each to appear in the merged object.
So, it will add a suffix (e.g., `.x`, `.y`) to each common variable to indicate which object (i.e., object `x` or object `y`) the variable came from, where object `x` is the first object—i.e., the object to which object `y` (the second object) is merged.
In general, apart from the keys, you should not include variables with the same name in two objects to be merged.
To prevent this, either remove or rename the shared variable in one of the objects, or include the shared variable as a key.
However, as described above, you should include it as a key ***only*** if you want to use its values to align the rows from each object.
Below is an example of merging two objects with the same variable name (i.e., points) that is not used as a key, using the `dplyr::full_join()` function.

@@

When two objects are merged that have different formats, the resulting data object inherits the format of the data object that has more levels of nesting.
For instance, consider that you want to merge two objects, object A and object B.
Object A is in `player` form and object B is in `player`-`season`-`week` form.
When you merge them, the resulting data object will be in `player`-`season`-`week` form.
Below, we merge data using the `dplyr::full_join()` function:

@@

A full outer join includes all rows in `x` **or** `y`.
It returns columns from `x` and `y`.
Here is how to merge two data frames using a full outer join (i.e., "full join"; `dplyr::full_join()`):

@@

A left outer join includes all rows in `x`.
It returns columns from `x` and `y`.
Here is how to merge two data frames using a left outer join ("left join"; `dplyr::left_join()`):

@@

A right outer join includes all rows in `y`.
It returns columns from `x` and `y`.
Here is how to merge two data frames using a right outer join ("right join"; `dplyr::right_join()`):

@@

An inner join includes all rows that are in **both** `x` **and** `y`.
An inner join will return one row of `x` for each matching row of `y`, and can duplicate values of records on either side (left or right) if `x` and `y` have more than one matching record.
It returns columns from `x` and `y`.
Here is how to merge two data frames using an inner join (`dplyr::inner_join()`):

@@

A semi join is a filter.
A left semi join returns all rows from `x` **with** a match in `y`.
That is, it filters out records from `x` that are not in `y`.
Unlike an [inner join](#innerJoin), a left semi join will never duplicate rows of `x`, and it includes columns from only `x` (not from `y`).
Here is how to merge two data frames using a left semi join (`dplyr::semi_join()`):

@@

An anti join is a filter.
A left anti join returns all rows from `x` **without** a match in `y`.
That is, it filters out records from `x` that are in `y`.
It returns columns from only `x` (not from `y`).
Here is how to merge two data frames using a left anti join (`dplyr::anti_join()`):

@@

A cross join combines each row in `x` with each row in `y` (`dplyr::cross_join()`).

@@

Below, we transform the data from long form to wide form.
The long data are in "`player`-`season`-`week` form".
We widen the data by two variables (`season` and `week`), using `tidyverse`, so that the data are now in "`player` form" (where each row is uniquely identified by the `ID` variable) using the `tidyr::pivot_wider()` function:

@@

Conversely, we can also restructure data from wide to long using the `tidyr::pivot_longer()` function.
Here are the data in long form, after they have been transformed from wide form using `tidyverse`:

### 2025-08-10 index.qmd
20250809 - label functions

# Automatically create a bib database for R packages
packagesForBibliography <- c(
  "petersenlab", "ffanalytics",
  "knitr", "rmarkdown", "broom", "broom.mixed", "psych", "downlit", "xml2", 
  "gsisdecoder", "progressr", "DescTools", "pwr", "pwrss", "WebPower", "XICOR", 
  "dagitty", "ggdag", "ggtext", "gghighlight", "ggExtra", "grid", "patchwork", 
  "pROC",  "lme4", "lmerTest", "MuMIn", "emmeans", "pbkrtest", "sjstats", 
  "AICcmodavg", "rstan", "brms", "tidybayes", "bbmle", "fitdistrplus", "sn", 
  "mclust", "magrittr", "viridis", "viridisLite", "msir", "plotly", "webshot2", 
  "quantmod", "fPortfolio", "NMOF", "nFactors", "xts", "zoo", "forecast",
  "stringi", "parallelly", "doParallel", "missRanger", "ggridges", "powerjoin", 
  "caret", "LongituRF", "gpboost", "cmdstanr", "mgcv", "rms", "car", "lavaan", 
  "lavaanPlot", "mice", "miceadds", "interactions", "robustbase", "ordinal", 
  "MASS", "data.table", "future", "future.apply", "SimDesign", "domir", 
  "GGally","Rglpk",
  "tidyverse", "ggplot2", "dplyr", "lubridate", # packages that are part of the tidyverse ecosystem
  "nflverse", "nflfastR", "nflseedR", "nfl4th", "nflreadr", "nflplotR", # packages that are part of the nflverse ecosystem
  "tidymodels", "rsample", "parsnip", "recipes", "tune", "yardstick", # packages that are part of the tidymodels ecosysyem
  "easystats", "see", "performance", "parameters", "correlation", "effectsize", # packages that are part of the easystats ecosystem
  "learnr", "remotes", "shiny", "PearsonDS", "lavaangui", "dynamic" # packages that are cited but not used for computation (and thus do not need to be installed by me or readers)
)

>>>>

# Automatically create a bib database for R packages
packagesForBibliography <- c(
  "petersenlab", "ffanalytics",
  "knitr", "rmarkdown", "broom", "broom.mixed", "psych", "downlit", "xml2", 
  "gsisdecoder", "progressr", "DescTools", "pwr", "pwrss", "WebPower", "XICOR", 
  "dagitty", "ggdag", "ggtext", "gghighlight", "ggExtra", "grid", "patchwork", 
  "pROC",  "lme4", "lmerTest", "MuMIn", "emmeans", "pbkrtest", "sjstats", 
  "AICcmodavg", "rstan", "brms", "tidybayes", "bbmle", "fitdistrplus", "sn", 
  "mclust", "magrittr", "viridis", "viridisLite", "msir", "plotly", "webshot2", 
  "quantmod", "fPortfolio", "NMOF", "nFactors", "xts", "zoo", "forecast",
  "stringi", "parallelly", "doParallel", "missRanger", "ggridges", "powerjoin", 
  "caret", "LongituRF", "gpboost", "cmdstanr", "mgcv", "rms", "car", "lavaan", 
  "lavaanPlot", "mice", "miceadds", "interactions", "robustbase", "ordinal", 
  "MASS", "data.table", "future", "future.apply", "SimDesign", "domir", 
  "GGally","Rglpk","TTR",
  "tidyverse", "ggplot2", "dplyr", "lubridate", # packages that are part of the tidyverse ecosystem
  "nflverse", "nflfastR", "nflseedR", "nfl4th", "nflreadr", "nflplotR", # packages that are part of the nflverse ecosystem
  "tidymodels", "rsample", "parsnip", "recipes", "tune", "yardstick", # packages that are part of the tidymodels ecosysyem
  "easystats", "see", "performance", "parameters", "correlation", "effectsize", # packages that are part of the easystats ecosystem
  "learnr", "remotes", "shiny", "PearsonDS", "lavaangui", "dynamic" # packages that are cited but not used for computation (and thus do not need to be installed by me or readers)
)

### 2025-08-10 machine-learning.qmd
20250809 - label functions

We set up the *k* folds using the `group_vfold_cv()` function of the `rsample` package [@R-rsample].

@@

We set up the LOO folds using the `loo_cv()` function of the `rsample` package [@R-rsample].

@@

Below, we fit a linear regression model with one predictor and evaluate it with [cross-validation](#sec-crossValidation).
We also evaluate its accuracy on the hold-out (test) data.
For each of the models, we fit and evaluate the models using the `tidymodels` ecosystem of packages [@R-tidymodels; @Kuhn2020_packages].
we specify our model formula using the `recipe()` function of the `recipes` package [@R-recipes].
We define the model using the `linear_reg()`, `set_engine()`, and `set_mode()` functions of the `parsnip` package [@R-parsnip].
We specify the workflow using the `workflow()`, `add_recipe()`, and `add_model()` functions of the `workflows` package [@R-workflows].
We fit the cross-validation model using the `fit_resamples()` function of the `tune` package [@R-tune].
We specify the accuracy metrics to evaluate using the `metric_set()` function of the `yardstick` package [@R-yardstick].
We fit the final model using the `fit()` function of the `workflows` package [@R-workflows].
We evaluate the accuracy of the model's predictions on the test data using the `accuracyOverall()` of the [`petersenlab`](https://cran.r-project.org/web/packages/petersenlab/index.html) package [@R-petersenlab].

@@

For the machine learning models, we perform the parameter tuning using the `tune()`, `tune_grid()`, `select_best()`, and `finalize_workflow()` functions of the `tune` package [@R-tune].
We specify the grid of possible parameter values using the `grid_regular()` function of the `dials` package [@R-dials].

@@

The random forest models were fit using the `ranger` package [@R-ranger; @Wright2017_packages].
We specify the grid of possible parameter values using the `grid_random()`, `update.parameters()`, `mtry()`, and `min_n()` functions of the `dials` package [@R-dials] and the `extract_parameter_set_dials()` function of the `hardhat` package [@R-hardhat].

@@

Approaches to estimating random forest models with longitudinal data are described by @Hu2023.
Below, we fit longitudinal random forest models using the `MERF()` function of the `LongituRF` package [@R-LongituRF].

@@

For identifying the optimal tuning parameters for boosting, we partition the training data into inner training data and validation data.
We randomly split the training data into 80% inner training data and 20% held-out validation data.
We then use the [mean absolute error](#sec-meanAbsoluteError) as our index of prediction accuracy on the held-out validation data.

@@

@@

```{r}
summary(gp_model_qb) # estimated random effects model

gp_model_qb_importance <- gpboost::gpb.importance(gp_model_fit_qb)
gp_model_qb_importance
```

>>>>

We set up the *k* folds using the `rsample::group_vfold_cv()` function of the `rsample` package [@R-rsample].

@@

We set up the LOO folds using the `rsample::loo_cv()` function of the `rsample` package [@R-rsample].

@@

Below, we fit a linear regression model with one predictor and evaluate it with [cross-validation](#sec-crossValidation).
We also evaluate its accuracy on the hold-out (test) data.
For each of the models, we fit and evaluate the models using the `tidymodels` ecosystem of packages [@R-tidymodels; @Kuhn2020_packages].
we specify our model formula using the `recipes::recipe()` function of the `recipes` package [@R-recipes].
We define the model using the `parsnip::linear_reg()`, `parsnip::set_engine()`, and `parsnip::set_mode()` functions of the `parsnip` package [@R-parsnip].
We specify the workflow using the `workflows::workflow()`, `workflows::add_recipe()`, and `workflows::add_model()` functions of the `workflows` package [@R-workflows].
We fit the cross-validation model using the `tune::fit_resamples()` function of the `tune` package [@R-tune].
We specify the accuracy metrics to evaluate using the `yardstick::metric_set()` function of the `yardstick` package [@R-yardstick].
We fit the final model using the `workflows::fit()` function of the `workflows` package [@R-workflows].
We evaluate the accuracy of the model's predictions on the test data using the `petersenlab::accuracyOverall()` of the [`petersenlab`](https://cran.r-project.org/web/packages/petersenlab/index.html) package [@R-petersenlab].

@@

For the machine learning models, we perform the parameter tuning using the `tune::tune()`, `tune::tune_grid()`, `tune::select_best()`, and `tune::finalize_workflow()` functions of the `tune` package [@R-tune].
We specify the grid of possible parameter values using the `dials::grid_regular()` function of the `dials` package [@R-dials].

@@

The random forest models were fit using the `ranger` package [@R-ranger; @Wright2017_packages].
We specify the grid of possible parameter values using the `dials::grid_random()`, `dials::update.parameters()`, `dials::mtry()`, and `dials::min_n()` functions of the `dials` package [@R-dials] and the `hardhat::extract_parameter_set_dials()` function of the `hardhat` package [@R-hardhat].

@@

Approaches to estimating random forest models with longitudinal data are described by @Hu2023.
Below, we fit longitudinal random forest models using the `LongituRF::MERF()` function of the `LongituRF` package [@R-LongituRF].

@@

For identifying the optimal tuning parameters for boosting, we partition the training data into inner training data and validation data.
We randomly split the training data into 80% inner training data and 20% held-out validation data.
We then use the [mean absolute error](#sec-meanAbsoluteError) as our index of prediction accuracy on the held-out validation data.
We use the `gpboost::gpb.Dataset()` function to specify the data, the `gpboost::GPModel()` function to specify the model, and the `gpboost::gpb.grid.search.tune.parameters()` function to identify the optimal tuning parameters for that model.

@@

We use the `gpboost::gpb.train()` function to fit the specified model with the specified tuning parameters and dataset.

@@

```{r}
summary(gp_model_qb) # estimated random effects model
```

We evaluate the importance of features using the `gpboost::gpb.importance()` and `gpboost::gpb.plot.importance()` functions:

```{r}
gp_model_qb_importance <- gpboost::gpb.importance(gp_model_fit_qb)
gp_model_qb_importance
```

### 2025-08-10 mixed-models.qmd
20250806 - robust load
20250809 - label functions

We compare nested models using an $F$-test with the `anova()` function.
We compare non-nested models using the `bbmle` [@R-bbmle] and `MuMIn` [@R-MuMIn] packages.

@@

We estimate the multilevel models using the `lmerTest` package [@R-lmerTest], which is an extension of the `lme4` package [@R-lme4; @Bates2015_packages]:

@@

To obtain standardized regression coefficients, we use the `standardize_parameters()` function of the `effectsize` package [@R-effectsize; @BenShachar2020_packages].
We use the `emmeans` package [@R-emmeans] to determine the model-implied number of points per season by position.

@@

To visualize the regression coefficients, we can use the `tidy()` function of the `broom.mixed` package [@R-broom.mixed], as in @fig-individualFantasyPointsRandomLinearFixedQuadraticSlopesInteractionCoefficients.

@@

load(url("https://osf.io/download/q6rjf/"))
```

@@

To visualize the regression coefficients, we can use the `tidy()` function of the `broom.mixed` package [@R-broom.mixed], as in @fig-bayesianRegressionCoefficients.

>>>>

We compare nested models using an $F$-test with the `stats::anova()` function.
We compare non-nested models using the `bbmle` [@R-bbmle] (`bbmle::AICtab()`) and `MuMIn` [@R-MuMIn] (`MuMIn::AICc()`) packages.

@@

We estimate the multilevel models using the `lmerTest::lmer()` function of the `lmerTest` package [@R-lmerTest], which is an extension of the `lme4` package [@R-lme4; @Bates2015_packages]:

@@

To obtain standardized regression coefficients, we use the `effectsize::standardize_parameters()` function of the `effectsize` package [@R-effectsize; @BenShachar2020_packages].
We use the `emmeans::emmeans()` function of the `emmeans` package [@R-emmeans] to determine the model-implied number of points per season by position.

@@

To visualize the regression coefficients, we can use the `broom.mixed::tidy()` function of the `broom.mixed` package [@R-broom.mixed], as in @fig-individualFantasyPointsRandomLinearFixedQuadraticSlopesInteractionCoefficients.

@@

petersenlab::robust_load("https://osf.io/download/q6rjf/") #load(url("https://osf.io/download/q6rjf/"))
```

@@

To visualize the regression coefficients, we can use the `broom.mixed::tidy()` function of the `broom.mixed` package [@R-broom.mixed], as in @fig-bayesianRegressionCoefficients.
